Disk Heating, Galactoseismology, and the Formation of Stellar Halos

### 2017-06-27 paper/paper.tex
section 2 start

The focus of this \documentname\ is on three such substructures just beyond the
historical ``end'' of the Galactic disk within the inner stellar halo.
Figure~\ref{fig:ting} \cite[reproduced with data from][]{li17}, shows the
spatial distribution of M giant stars associated with these three
substructures: the so-called Monoceros Ring (Mon; also known as the Giant
Anticenter Stellar Structure, GASS), the Triangulum-Andromeda clouds (TriAnd),
and A13.
Each of these substructures were originally identified as over-densities in
stellar number counts relative to the expected global structure of the disk or
inner stellar halo.\footnote{Here we show just M giant stars that have been
previously identified as candidate members of the structures, however, some of
the structures have also been detected in MSTO stars.}
The same region of the sky has been shown to be richly structured on even
smaller scales \cite{slater14,martin14,deason14}, but here we consider only the
larger structures.
Unlike most stellar streams, Mon/GASS, TriAnd, and A13 are present at a range
of low to moderate Galactic latitudes and span large areas of the sky --- we
will hereafter refer to them collectively as the ``low-latitude structures''.
The basic properties of the low-latitude structures are summarized below:
\begin{description}
\item{\it Mon/GASS} is an arc-like or partial-ring feature of stars beyond the previously-expected edge of the Galactic disk, $\approx 5~\kpc$ beyond the Sun in cylindrical radius \cite{robin92}.
Stars attributed to this feature span a large area of the sky and large range
in distance: Galactic longitudes $\approx 120^\circ \lesssim l \lesssim
240^\circ$, latitudes $-30^\circ \lesssim b \lesssim +40^\circ$, and
Heliocentric distances $5\lesssim d_\odot \lesssim 10~\kpc$
\cite{Morganson:2016}.
Radial velocity measurements of M giant stars associated with the structure
follow a clear trend in mean velocity with Galactic longitude and have a
velocity dispersion much smaller than the stellar halo \cite{crane03}.
\item{\it TriAnd} was first discovered as a diffuse over-density of M giant
stars covering the area $\approx 100^\circ \lesssim l \lesssim 150^\circ$,
$\approx 20^\circ \lesssim b \lesssim 40^\circ$, overlapping the Mon/GASS
structure on the sky but at larger Heliocentric distances of $\approx
15$--$25~\kpc$ \cite{rochapinto04}.
The M giants again exhibit a coherent radial-velocity sequence with a
dispersion much smaller than the halo \cite{rochapinto04}.
Deep photometry in the region revealed MSTO stars associated with the structure
and proposed the existence of a second main sequence (``TriAnd 2'') at larger
distance, $\approx 30$--$35~\kpc$ \cite{martin07,martin14}.
\item{\it A13} is another tenuous association of M giants in the north Galactic
hemisphere in the area $\approx 125^\circ \lesssim l \lesssim 210^\circ$,
$\approx 20^\circ \lesssim b \lesssim 40^\circ$ at approximate distances of
$\approx 12$--$20~\kpc$.
It was initially discovered by applying a group finding algorithm
\cite{sharma09} to all M giants in the 2MASS photometric catalogue
\cite{sharma10}.
Again, radial velocities of M giants in this structure have a small velocity
dispersion around a roughly linear trend with Galactic longitude \cite{li17}
\end{description}

@@

From clustering in positions or distance alone, many candidate groups and
over-densities of M giants have been identified in the outer disk / inner halo.
Over the last five years our group has pursued a spectroscopic campaign to
follow up members of these candidate structure with the aims of (1) confirming
the existence of substructure in velocities, (2) measuring chemical abundances,
and (3) studying the constituent stellar populations.
These goals then inform our own efforts to produce plausible dynamical
formation scenarios using simulations.
We were interested in particular in {\it avoiding} the collimated stellar streams that have been well-studied in prior work \cite[such as Sgr, Orphan, GD1 and Pal 5 --- see, e.g.,][]{law10,koposov10,kuepper15,bovy16}
and instead considered groups that appeared as diffuse, amorphous, extended stellar structures such as TriAnd 1 and 2 \cite{rochapinto04} and A13 \cite{sharma10}.
Initial interpretations of these morphologies suggested the structures could be {\it shells} --- debris from the disruption of satellite galaxies on near-radial orbits \cite{johnston08} --- as seen from an internal perspective.

In our first results \cite{sheffield14}, we extended a prior sample of TriAnd
M giants \cite{rochapinto04} by obtaining spectra of all candidate M giants identified by applying color-magnitude cuts to stars in the TriAnd region of the sky.
We found that the M giants occupied two distinct loci in color-magnitude space, which we associated with the two MSTO identified in optical data by \cite{martin07}, named TriAnd 1 and TriAnd 2 respectively.
Stars in both TriAnd 1 and TriAnd 2  form  clear over-densities in radial velocity with low dispersion ($\sim 25~\kms$) compared to the background halo velocity distribution.
They both follow the same  sequence in velocity with a steady gradient of mean Galactic standard-of-rest (GSR) radial velocity ($v_{\rm GSR}$) from positive to negative $v_{\rm GSR}$ with increasing $l$.
In this work, we presented a model that simultaneously and approximately reproduces TriAnd 1 and 2 as tidal debris stripped over two separate pericentric passages from a single accreted satellite on a low-eccentricity, retrograde, near-planar orbit.
The debris structures in the simulations were morphologically closer to {\it streams} rather
than {\it shells}, but still subtended large areas on the sky as observed from
the Sun's position.

In subsequent work we continued the spectroscopic survey to look at A13, the new group M giant discovered through the group-finding algorithm \cite{sharma10}. A13 overlaps the TriAnd clouds in Galactic longitude at one end and GASS at the other end, and is apparent in the Northern Galactic hemisphere at slightly brighter magnitudes than TriAnd. The spectra show that, like the TriAnd clouds, this group has a coherent velocity structure with low dispersion and a steady gradient with longitude, $l$, confirming the genuine association of its members \cite{li17}.

@@

Figure~\ref{fig:ting_vel} (reproduced from \cite{li17}), summarizes the line-of-sight (GSR) velocity trends of M giants in each of the three low-latitude structures.
Not only do these groups all have low dispersions relative to the stellar halo
--- which confirms  groups themselves as real --- they also
appear to {\it collectively} exhibit a continuous gradient with Galactic longitude,
$l$.
This latter finding suggests that the groups may also be associated with each other.

>>>>

The focus of this \documentname\ is on three such substructures just beyond the
historical ``end'' of the Galactic disk within the inner stellar halo.
Figure~\ref{fig:ting} \cite[reproduced with data from previous work, ][]{li17},
shows the spatial distribution of M giant stars associated with these three
substructures: the so-called Monoceros Ring (Mon; also known as the Giant
Anticenter Stellar Structure, GASS), the Triangulum-Andromeda clouds (TriAnd),
and A13.
Each of these substructures were originally identified as over-densities in
stellar number counts relative to the expected global structure of the disk or
inner stellar halo.\footnote{Here we show just M giant stars that have been
previously identified as candidate members of the structures, however, some of
the structures have also been detected in MSTO stars.}
The same region of the sky has been shown to be richly structured on even
smaller scales \cite{slater14,martin14,deason14}, but here we consider only the
larger structures.
Unlike most stellar streams, Mon/GASS, TriAnd, and A13 are present at a range
of low to moderate Galactic latitudes and span large areas of the sky --- we
will hereafter refer to them collectively as the ``low-latitude structures''.
The basic properties of the low-latitude structures are summarized below:
\begin{description}
\item{\it Mon/GASS} is an arc-like or partial-ring feature of stars beyond the previously-expected edge of the Galactic disk, $\approx 5~\kpc$ beyond the Sun in cylindrical radius \cite{robin92}.
Stars attributed to this feature span a large area of the sky and large range
in distance: Galactic longitudes $\approx 120^\circ \lesssim l \lesssim
240^\circ$, latitudes $-30^\circ \lesssim b \lesssim +40^\circ$, and
Heliocentric distances $5\lesssim d_\odot \lesssim 10~\kpc$
\cite{Morganson:2016}.
Radial velocity measurements of M giant stars associated with the structure
follow a clear trend in mean velocity with Galactic longitude and have a
velocity dispersion much smaller than the stellar halo \cite{crane03}.
\item{\it TriAnd} was first discovered as a diffuse over-density of M giant
stars covering the area $\approx 100^\circ \lesssim l \lesssim 150^\circ$,
$\approx 20^\circ \lesssim b \lesssim 40^\circ$, overlapping the Mon/GASS
structure on the sky but at larger Heliocentric distances of $\approx
15$--$25~\kpc$ \cite{rochapinto04}.
The M giants again exhibit a coherent radial-velocity sequence with a
dispersion much smaller than the halo \cite{rochapinto04}.
Deep photometry in the region revealed MSTO stars associated with the structure
and proposed the existence of a second main sequence (``TriAnd 2'') at larger
distance, $\approx 30$--$35~\kpc$ \cite{martin07,martin14}.
\item{\it A13} is another tenuous association of M giants in the north Galactic
hemisphere in the area $\approx 125^\circ \lesssim l \lesssim 210^\circ$,
$\approx 20^\circ \lesssim b \lesssim 40^\circ$ at approximate distances of
$\approx 12$--$20~\kpc$.
It was initially discovered by applying a group finding algorithm
\cite{sharma09} to all M giants in the 2MASS photometric catalogue
\cite{sharma10}.
Again, radial velocities of M giants in this structure have a small velocity
dispersion around a roughly linear trend with Galactic longitude \cite{li17}
\end{description}

@@

From clustering in positions or distance alone, many candidate groups and
over-densities of M giants have been identified in the outer disk / inner halo.
Over the last five years our group has pursued a spectroscopic campaign to
follow up members of these candidate structure with the aims of (1) confirming
the existence of substructure in velocities, (2) measuring chemical abundances,
and (3) studying the constituent stellar populations.
These goals then inform our own efforts to produce plausible dynamical
formation scenarios using simulations.
In particular, we {\it avoid} the collimated stellar streams that have been well-studied in prior work \cite[such as Sgr, Orphan, GD1 and Pal 5 --- see, e.g.,][]{law10,koposov10,kuepper15,bovy16}
and instead focus on groups that appear as diffuse, amorphous, extended stellar structures such as TriAnd \cite{rochapinto04} and A13 \cite{sharma10}.
Initial interpretations of these morphologies suggested the structures could be {\it shells} --- debris from the disruption of satellite galaxies on near-radial orbits \cite{johnston08} --- as seen from an internal perspective.

In our first results, we extended a prior sample of TriAnd M giants
\cite{rochapinto04} by obtaining spectra of all candidate M giants identified
by applying color-magnitude cuts to stars in the TriAnd region of the sky
\cite{sheffield14}.
We identified M giants associated with the two proposed MSTO TriAnd structures
\cite[TriAnd 1 and 2, as named by][]{martin07,martin14}.
M giant stars in both TriAnd 1 and 2 form clear over-densities in radial
velocities with a small dispersion, $\sigma_v \approx 25~\kms$, compared to the
background halo velocity distribution, $\sigma_v \approx 100~\kms$.
The radial velocities of M giants in both TriAnd 1 and 2 follow the same
sequence in velocity with a steady gradient of mean Galactic standard-of-rest
(GSR) radial velocity ($v_{\rm GSR}$) from positive to negative $v_{\rm GSR}$
with increasing Galactic longitude, $l$.
We presented a model that simultaneously and approximately reproduces TriAnd 1
and 2 as tidal debris stripped over two separate pericentric passages from a
single accreted satellite on a low-eccentricity, retrograde, near-planar orbit.
The debris structures in the simulations were morphologically closer to {\it
streams} rather than {\it shells}, but still subtended large areas on the sky
as observed from the Sun's position.
We note, however, that because of large distance uncertainties, the M giants in
TriAnd 1 and 2 are indistinguishable and overlap in distance, velocity
distribution, and sky position; the existence of two distinct structures rather
than a single extended structure has yet to be conclusively demonstrated.
Hereafter, we therefore refer to the TriAnd structures collectively, rather
than individually.

In subsequent work, we continued this spectroscopic survey by observing M giant
stars in A13 \cite{sharma10}.
A13 overlaps the TriAnd clouds in Galactic longitude at one end, and GASS at
the other end, but is apparent in the Northern Galactic hemisphere at slightly
brighter magnitudes than TriAnd.
The spectra show that, like the TriAnd clouds, this group has a
coherent velocity structure with low dispersion and a steady gradient with
longitude, $l$, confirming the genuine association of its members \cite{li17}.

@@

% APW COMMENT: left off here, but no mention of GASS above - figure shows all 3
% structures. Should reference Crane et al. study of GASS?

Figure~\ref{fig:ting_vel} \cite[reproduced with data from previous work,][]{li17}, summarizes the line-of-sight (GSR) velocity trends of M giants in each of the three low-latitude structures.
Not only do these groups all have low dispersions relative to the stellar halo
--- which confirms  groups themselves as real --- they also
appear to {\it collectively} exhibit a continuous gradient with Galactic longitude,
$l$.
This latter finding suggests that the groups may also be associated with each other.
